%*******************************************************
% Acknowledgments
%*******************************************************
\pdfbookmark[1]{Acknowledgments}{acknowledgments}

%\begin{flushright}{\slshape
%    We have seen that computer programming is an art, \\
%    because it applies accumulated knowledge to the world, \\
%    because it requires skill and ingenuity, and especially \\
%    because it produces objects of beauty.} \\ \medskip
%    --- \defcitealias{knuth:1974}{Donald E. Knuth}\citetalias{knuth:1974} \citep{knuth:1974}
%\end{flushright}



\bigskip

\begingroup
\let\clearpage\relax
\let\cleardoublepage\relax
\let\cleardoublepage\relax
\chapter*{Acknowledgements}
I want to thank my supervisor Matteo Smerlak for his support, orientation, apprehension and unlimited tolerance during this challenging exercise. \textit{Merci bien pour cette immense opportunité.}
Special gratitude to Prof$.$ Dr$.$ Peter F. Stadler for reading through my first draft and supporting this thesis submission. 

Thanks to Ian Hatton's patience and unlimited comments, that significantly improve this essay. A special thank goes to my collaborator, colleague and friend Opuu Vaitea who contributed to the publications part of this thesis. 

I thank the Structure of Evolution Group at the Max Planck Institute for Mathematics in the Sciences, especially Vincent Messow and Camila Bräutigam, for valuable discussions and insightful comments. 

My warmest thanks extend to all MPI-MIS administration staff for their supports in making my stay in Leipzig comfortable and conducive to learning. Thanks to the AIMS network for their help and time to share great and interesting information and knowledge. 

Many thanks also to Sayan Mukherjee for ensuring that my career starts as this thesis ends. I am looking forward to working with you!

I am also particularly grateful to my friends Abdulrraouf Biala and Josephina Burger for their unlimited emotional and social supports throughout my stay in Leipzig. 

I would also like to thank my Leipziger people, Saida, Saidou, Anne, Hamza,  Taba and Mokobe for their familiarity, encouragement and social support. 

Thanks to Louisa Kienzler for the unique moments and valuable comments that contributed to improving this work. 

Thanks to the LEC church family for their spiritual and social supports, especially Verona Hivemuine Black, for being there precisely when I needed it.

\textit{Et bien sûr, merci à ma famille et Ousmanou Djika pour leur soutien inconditionel. }
%\textit{Je vous aime de tout mon coeur...}

%Regarding the typography and other help, many thanks go to Marco
%Kuhlmann, Philipp Lehman, Lothar Schlesier, Jim Young, Lorenzo
%Pantieri and Enrico Gregorio\footnote{Members of GuIT (Gruppo
%Italiano Utilizzatori di \TeX\ e \LaTeX )}, J\"org Sommer,
%Joachim K\"ostler, Daniel Gottschlag, Denis Aydin, Paride
%Legovini, Steffen Prochnow, Nicolas Repp, Hinrich Harms,
%Roland Winkler, Jörg Weber, Henri Menke, Claus Lahiri,
%Clemens Niederberger, Stefano Bragaglia, Jörn Hees,
%Scott Lowe, Dave Howcroft, Jos\'e M. Alcaide, David Carlisle,
%Ulrike Fischer, Hugues de Lassus, Csaba Hajdu, Dave Howcroft, 
%and the whole \LaTeX-community for support, ideas and
%some great software.

\bigskip

%\noindent\emph{Regarding \mLyX}: The \mLyX\ port was intially done by
%\emph{Nicholas Mariette} in March 2009 and continued by
%\emph{Ivo Pletikosi\'c} in 2011. Thank you very much for your
%work and for the contributions to the original style.


\endgroup
